rang: Reconstructing reproducible R computational environments

### 2023-05-18 paper/paper.qmd
Minor update on the paper

1. By default, the Docker image building process downloads source packages from their sources and then compiles them. This step depends on the future availability of R packages on CRAN (which is extremely likely to be the case in the near future, given the continuous availability since 1997-04-23) [^1], Bioconductor, and Github. However, it is also possible to cache (or archive) the source packages now. The archived R packages can then be used during the building process instead. The significance of this step in terms of long-term computational reproducibility will be discussed in Section 4.

@@

# Preparing research compendia with long-term computational reproducibility

The above six examples show how powerful `rang` is to reconstruct tricky computational environments which have not been completely declared in the literature. Although we position `rang` mostly as an archaeological tool, we think that `rang` can also be used to prepare research compendia of current research. We can't predict the future but research compendia generated by `rang` would probably have long-term computational reproducibility.

@@

This paper presents `rang`, a solution to (re)construct R computational environments based on Docker. As the six examples in Section 3 show, `rang` can be used archaeologically to rerun old code, many of them not executable without the analytic and reconstruction processes facilitated by `rang`. These retrospective use cases demonstrate how versatile `rang` is. `rang` is also helpful for prospective usage, as demonstrated in Section 4 where an executable compendium is introduced.

@@

Moreover, the (re)construction process based on `rang` takes notably more time than other solutions because all packages are compiled from source. `rang` trades computational efficiency of this often one-off (re)constructing process for correctness, backward compatibility and independence from any commercial backups of software repositories such as MRAN. In the Vignette of `rang` ([https://cran.r-project.org/web/packages/rang/vignettes/faq.html](https://cran.r-project.org/web/packages/rang/vignettes/faq.html)), we list all limitations: (1) `rang` does not support R < 1.3.1, (2) `rang` can generate Ubuntu/Debian-based Docker images and therefore non-Linux features (e.g. WinBUGS) are not supported, (3) `dockerize()` cannot cache OS-level packages (`deb` packages), (4) R packages with changing system requirements might be incorrectly queried, (5) caching of R packages and R source must be enforced when the R version is too old, (6) R packages on CRAN, Bioconductor, and Github, as well as Rocker images, might not be available in the future. All caveats listed have no solution, except point (6) can be mitigated by backing up the generated Docker image (see the section on research compendia).

| Function           | Purpose                                                                              |
|--------------------|--------------------------------------------------------------------------------------|
| `as_pkgrefs`       | Convert data structures (e.g. `sessionInfo`) into package references for `resolve()` |
| `convert_edgelist` | Convert the resolved results to edgelist                                             |
| `dockerize`        | Dockerize the resolved result                                                        |
| `export_rang`      | Export the resolved result as installation script                                    |
| `export_renv`      | Export the resolved result as renv lockfile                                          |
| `query_sysreqs`    | Query for system requirements                                                        |
| `resolve`          | Resolve dependencies of R packages                                                   |

>>>>

1. By default, the Docker image building process downloads source packages from their sources and then compiles them. This step depends on the future availability of R packages on CRAN (which is extremely likely to be the case in the near future, given the continuous availability since 1997-04-23) [^1], Bioconductor, and Github. However, it is also possible to cache (or archive) the source packages now. The archived R packages can then be used during the building process instead. The significance of this step in terms of long-term computational reproducibility will be discussed in the section on executable compendia.

@@

# Preparing executable compendia with long-term computational reproducibility

The above six examples show how powerful `rang` is to reconstruct tricky computational environments which have not been completely declared in the literature. Although we position `rang` mostly as an archaeological tool, we think that `rang` can also be used to prepare executable research compendia of current research. We cannot predict the future but research compendia generated by `rang` would probably have long-term computational reproducibility.

@@

This paper presents `rang`, a solution to (re)construct R computational environments based on Docker. As the six examples in Section 3 show, `rang` can be used archaeologically to rerun old code, many of them not executable without the analytic and reconstruction processes facilitated by `rang`. These retrospective use cases demonstrate how versatile `rang` is. `rang` is also helpful for prospective usage, as demonstrated in the section where an executable compendium is introduced.

@@

Moreover, the (re)construction process based on `rang` takes notably more time than other solutions because all packages are compiled from source. `rang` trades computational efficiency of this often one-off (re)constructing process for correctness, backward compatibility and independence from any commercial backups of software repositories such as MRAN. In the Vignette of `rang` ([https://cran.r-project.org/web/packages/rang/vignettes/faq.html](https://cran.r-project.org/web/packages/rang/vignettes/faq.html)), we list all limitations: (1) `rang` does not support R < 1.3.1, (2) `rang` can generate Ubuntu/Debian-based Docker images and therefore non-Linux features (e.g. WinBUGS) are not supported, (3) `dockerize()` cannot cache OS-level packages (`deb` packages), (4) R packages with changing system requirements might be incorrectly queried, (5) caching of R packages and R source must be enforced when the R version is too old, (6) R packages on CRAN, Bioconductor, and Github, as well as Rocker images, might not be available in the future. All caveats listed have no solution, except point (6) can be mitigated by backing up the generated Docker image (see the section on executable compendia).

| Function           | Purpose                                                              |
|--------------------|----------------------------------------------------------------------|
| `as_pkgrefs`       | Convert data structures (e.g. `sessionInfo`) into package references |
| `convert_edgelist` | Convert the resolved results to edgelist                             |
| `dockerize`        | Dockerize the resolved result                                        |
| `export_rang`      | Export the resolved result as installation script                    |
| `export_renv`      | Export the resolved result as renv lockfile                          |
| `query_sysreqs`    | Query for system requirements                                        |
| `resolve`          | Resolve dependencies of R packages                                   |
